Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-25 _tex/index.tex
Built site for gh-pages

However, since the discriminal process of a single stimulus is not
directly observable, the \emph{law of comparative judgment} becomes
essential. This law states that in pairwise comparisons, the stimulus
positioned further along the trait continuum is perceived as having a
higher level of that trait. Thus, the theory assumes the observed
dichotomous outcome is determined by the distribution of the difference
between the underlying discriminal processes of the stimuli, referred to
as the \emph{discriminal difference}. This indicates that the outcome
depends on the relative distance between stimuli, rather than their
absolute positions on the trait continuum.

These concepts are more easily understood through an example. For
instance, in the context of evaluating text quality,
Figure~\ref{fig-discriminal_process} could depict the underlying
discriminal process distributions for two written texts, highlighting
differences in their discriminal dispersions and modal discriminal
processes along the quality trait continuum. Furthermore,
Figure~\ref{fig-discriminal_difference} could display the discriminal
difference distribution for these texts, showing that text A is
perceived to exhibit significantly higher quality than text B, as
indicated by the shaded gray area. Consequently, the dichotomous outcome
of this comparison would probably favor text A.

@@

Importantly, the general form of Thurstone's theory primarily addressed
pairwise comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927b}. Thus, for practical application,
Thurstone introduced five distinct cases derived from this general form,
each defined by progressively simplifying assumptions.
Table~\ref{tbl-thurstone_cases} summarizes these cases, highlighting key
assumptions such as the distribution of discriminal processes, the
similarity of discriminal dispersions across stimuli, the correlation
between stimuli, and the number of judges performing the comparisons.
For a comprehensive discussion of this progression, refer to
\citet{Thurstone_1927b} and \citet[pp.~248-253]{Bramley_2008}.

@@

Despite its reliance on the largest number of simplifying assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
Case V remains the most widely used case in the CJ literature. This
popularity is largely due to its simplified statistical representation
in the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
assumptions of Case V, with one key difference: while Case V assumes a
Normal distribution for the discriminal processes of stimuli, the BTL
model uses the more mathematically tractable Logistic distribution
\citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). This substitution has little impact on
the model's estimation or interpretation, as the Normal and Logistic
distributions share similar statistical properties, differing only by a
scaling factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I} (see
Figure~\ref{fig-logistic_vs_normal}).

@@

The assumption of \emph{zero correlation between stimuli} can be better
understood through an example. For instance, when using pairwise
comparisons to evaluate text quality, the assumption implies that a
judge's perception of a trait in one text does not influence his
perception of the same trait in another text. Thurstone attributed this
independence to the cancellation of potential judges' biases, driven by
two opposing and equally weighted effects occurring during the pairwise
comparisons \citep[pp.~268]{Thurstone_1927b}. This cancellation was
mathematically demonstrated by \citet{Andrich_1978}, using the BTL model
under the assumption of additive biases in the discriminal processes.
However, it is easy to imagine at least two scenarios where the zero
correlation assumption almost certainly does not hold: when the pairwise
comparison involves multidimensional, complex traits with heterogeneous
stimuli, and when an additional hierarchical structure is relevant to
the stimuli.

In the first scenario, the intricate aspects of multidimensional,
complex traits and heterogeneous stimuli may introduce dependencies
between stimuli. Research on text quality suggests that when judges
evaluate these traits, they often rely on various intricate aspects of
the stimuli to form their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. In
this context, it is not inconceivable that these aspects, being neither
equally weighted nor opposing, may unevenly influence judges'
perceptions, resulting in biases that resist cancellation. For example,
this might occur when a judge assessing the argumentative quality of a
text places disproportionate emphasis on grammatical accuracy,
ultimately favoring texts with fewer errors but weaker arguments. While
direct evidence for this specific scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of judges' biases,
supporting the idea that the different factors influencing pairwise
comparisons may not always cancel out.

In the second scenario, the shared context or inherent connections
created by the additional hierarchical structure may introduce
dependencies between stimuli, a statistical phenomenon commonly known as
clustering \citep{Everitt_et_al_2010}. Nevertheless, despite recognizing
such hierarchical structures in CJ data, the statistical handling of
this extra source of dependency in the CJ literature has been
inadequate. For instance, in cases where the CJ data included multiple
samples of stimuli from the same individuals, researchers have often
relied on (averaged) estimated BTL scores to conduct subsequent analyses
and tests at the individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
This approach, however, has the significant limitation of ignoring the
uncertainty associated with the scores (refer to section
Section~\ref{sec-theory-issue2} for a detailed discussion of this
issue).

In contrast, the assumption of \emph{equal dispersion between stimuli}
suggests that the variability in the perceived trait level of the
stimuli is the same across stimuli. While Thurstone acknowledged that
this assumption may be violated when ``dealing with less conspicuous
attributes or with less homogeneous stimuli''
\citep[pp.~374]{Thurstone_1927a}, no study explicitly proposes that this
assumption could also be violated due to the presence of an additional
hierarchical (grouping) structure relevant to the texts. One such
scenario might arise, for example, when comparing texts produced by
university and secondary school students. In this case, university
students may consistently (or more precisely) produce higher-quality
texts, while secondary school students, who exhibit a broader range of
writing abilities, would show greater variability in the quality of
their texts. Although this example is somewhat contrived, it effectively
illustrates how assuming equal dispersions across texts can overlook
meaningful differences in the reliability of text quality across groups
or individuals.

{``however'' paragraph}

>>>>

Nevertheless, the theory posits that because the discriminal process of
a single stimulus is not directly observable, the \emph{law of
comparative judgment} becomes essential. This law states that in
pairwise comparisons, the stimulus positioned further along the trait
continuum is perceived as having a higher level of that trait. This
emphasize that the outcome depends on the relative distance between
stimuli, rather than their absolute positions on the trait continuum.
Thus, the theory assumes the observed dichotomous outcome is determined
by the distribution of the difference between the underlying discriminal
processes of the stimuli, referred to as the \emph{discriminal
difference}.

These concepts are more easily understood through an example. For
instance, in the context of evaluating text quality,
Figure~\ref{fig-discriminal_process} depicts the underlying discriminal
process distributions for two written texts. The figure highlights
differences in the texts' positions along the quality trait continuum,
represented by their modal discriminal processes, as well as differences
in the variability of their perceived trait levels, represented by their
discriminal dispersions. Furthermore,
Figure~\ref{fig-discriminal_difference} displays the discriminal
difference distribution for these texts. This figure shows that text A
is perceived to exhibit significantly higher quality than text B, as
indicated by the shaded gray area. Consequently, the dichotomous outcome
of this comparison would likely favor text A.

@@

Importantly, the general form of Thurstone's theory primarily addressed
pairwise comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927b}. Thus, to facilitate its practical
application, Thurstone developed five distinct cases derived from this
general form. Each case progressively introduces additional simplifying
assumptions related to different features of the theory, such as the
distribution of discriminal processes, the uniformity of discriminal
dispersions across stimuli, the correlation between stimuli, and the
number of judges performing the comparisons. A summary of these cases is
provided in Table~\ref{tbl-thurstone_cases}. For a detailed examination
of this progression, refer to \citet{Thurstone_1927b} and
\citet[pp.~248-253]{Bramley_2008}.

@@

Despite its reliance on the largest number of simplifying assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
Case V remains the most widely used case in the CJ literature. This
popularity is largely due to its simplified statistical representation
in the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
assumptions of Case V, with one key difference: while Case V assumes a
Normal distribution for the discriminal processes of the stimuli, the
BTL model uses the more mathematically tractable Logistic distribution
\citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). This substitution has little impact on
the model's estimation or interpretation, as the Normal and Logistic
distributions share similar statistical properties, differing only by a
scaling factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I} (see
Figure~\ref{fig-logistic_vs_normal}).

@@

The assumption of \emph{zero correlation between stimuli} can be better
understood through an example. For instance, when using pairwise
comparisons to evaluate text quality, the assumption implies that a
judge's perception of a trait in one text does not influence his
perception of the same trait in another text. Thurstone attributed this
independence to the cancellation of potential judges' biases, driven by
two opposing and equally weighted effects occurring during the pairwise
comparisons \citep[pp.~268]{Thurstone_1927b}. This cancellation was
mathematically demonstrated by \citet{Andrich_1978}, using the BTL model
under the assumption of discriminal processes with additive biases.
However, it is easy to imagine at least two scenarios where the zero
correlation assumption almost certainly does not hold: when the pairwise
comparison involves multidimensional, complex traits with heterogeneous
stimuli, and when an additional hierarchical structure is relevant to
the stimuli.

In the first scenario, the intricate aspects of multidimensional,
complex traits may introduce dependencies between heterogeneous stimuli
due to certain judges' biases that resist cancellation. Research on text
quality indicates that when judges evaluate such traits, they often rely
on various intricate aspects of the stimuli to form their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. In
this context, it is not inconceivable that these aspects, being neither
equally weighted nor opposing, may unevenly influence judges'
perceptions, resulting in biases that resist cancellation. For example,
this could occur when a judge assessing the argumentative quality of a
text places disproportionate emphasis on grammatical accuracy,
ultimately favoring texts with fewer errors but weaker arguments. While
direct evidence for this specific scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of judges' biases,
supporting the idea that the factors influencing pairwise comparisons
may not always cancel out.

In the second scenario, the shared context or inherent connections
created by the additional hierarchical structure may introduce
dependencies between stimuli, a statistical phenomenon commonly known as
clustering \citep{Everitt_et_al_2010}. Nevertheless, despite recognizing
such hierarchical structures in CJ data, the statistical handling of
this extra source of dependency in the CJ literature has been
inadequate. For instance, in cases where the CJ data included multiple
samples of stimuli from the same individuals, researchers have often
relied on (averaged) estimated BTL scores to conduct subsequent analyses
and tests at the individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
This approach, however, has the significant limitation of ignoring the
uncertainty associated with the BTL scores (refer to section
Section~\ref{sec-theory-issue2} for a detailed discussion of this
issue).

In contrast, the assumption of \emph{equal dispersion between stimuli}
suggests that the variability in the perceived trait level of the
stimulus is the same across all stimuli. While Thurstone acknowledged
that this assumption may be violated when ``dealing with less
conspicuous attributes or with less homogeneous stimuli''
\citep[pp.~374]{Thurstone_1927a}, no study explicitly proposes that this
assumption could also be violated due to the presence of an additional
hierarchical (grouping) structure relevant to the texts. One such
scenario might arise, for example, when comparing texts produced by
university and secondary school students. In this case, university
students may consistently (or more precisely) produce higher-quality
texts, while secondary school students, who exhibit a broader range of
writing abilities, would show greater variability in the quality of
their texts. Although this example is somewhat contrived, it effectively
illustrates how assuming equal dispersions across texts can overlook
meaningful differences in the reliability of text quality across groups
or individuals.

{``However'' paragraph}
